---
title: "Spatial Econometrics"
subtitle: "Spatial Point Patterns"
author: "Barcaroli Clement, Gonzalez-Garcia Francisco, Mondelli Martin, Senseby Dorian"
date: "`r Sys.Date()`"
format:
  pdf:
    toc: true
    toc-title: "Table of Contents"
    toc-depth: 3
    number-sections: true
    colorlinks: true
    pdf-engine: lualatex
    documentclass: scrartcl   # or "article" / "report" if you prefer
    classoption: [a4paper, 11pt]
    geometry: 
      - margin=1in
    code-fold: true
    code-summary: "Show code"
    highlight-style: tango
execute:
  warning: false
  message: false
editor:
  visual:
    wrap: 72
---

```{r setup, include=FALSE}

#Libraries
if (!requireNamespace("sf", quietly = TRUE))          install.packages("sf")
if (!requireNamespace("mapview", quietly = TRUE))     install.packages("mapview")
if (!requireNamespace("rnaturalearth", quietly = TRUE)) install.packages("rnaturalearth")
if (!requireNamespace("spatstat", quietly = TRUE))    install.packages("spatstat")
if (!requireNamespace("leaflet", quietly = TRUE))     install.packages("leaflet")
if (!requireNamespace("raster", quietly = TRUE))      install.packages("raster")
if (!requireNamespace("terra", quietly = TRUE))       install.packages("terra")
if (!requireNamespace("readxl", quietly = TRUE))      install.packages("readxl")
if (!requireNamespace("knitr", quietly = TRUE))       install.packages("knitr")
if (!requireNamespace("fields", quietly = TRUE))      install.packages("fields")
if (!requireNamespace("scales", quietly = TRUE))      install.packages("scales")

# Load the libraries
library(sf)
library(mapview)
library(rnaturalearth)
library(spatstat)
library(spatstat.geom)
library(leaflet)
library(raster)
library(terra)
library(readxl)
library(tidyverse)
library(fields)
library(scales)
library(knitr)
library(sp)
```

```{r}
knitr::opts_chunk$set(
  echo = TRUE,       # afficher le code par défaut
  results = "hold",  # tous les outputs ensemble
  fig.show = "hold", # tous les graphiques ensemble
  warning = FALSE,
  message = FALSE
)

```

Importing main data

```{r}
# Adaptative path to data
path_data = paste(getwd(), "2025-07-metropolitan-outcomes.csv", sep = "/")

police = read.csv(path_data)

```

# Data base presentation

For this project we used two datasets:

-   For the regressions we used a dataset containing information on the population and GDP among other variables that we did not use in this project (<https://www.ons.gov.uk/economy/grossdomesticproductgdp/datasets/regionalgrossdomesticproductlocalauthorities>)

-   or testing CSR, we used the July 2025 street-level crime and outcomes data provided by [UK Police]{.underline}, which contains geocoded crime events with latitude and longitude coordinates. <https://data.police.uk/data/archive/2025-07.zip>

# I/ Spatial Point Patterns Introduction

Counting number of NAs

```{r}
sum(is.na(police$Longitude))
sum(is.na(police$Latitude))
```

Filtering and we keep only police_clean

```{r}
police_clean <- police[!is.na(police$Longitude) & !is.na(police$Latitude),]
```

Making it sf

```{r}
my_coords <- st_as_sf(police_clean, coords = c("Longitude", "Latitude"), crs = 4326)
my_coords <- st_transform(my_coords, 27700)
```

Plot it

```{r}
mapview(my_coords) #If it doesn't work try with leaflet
```

```{r}
coords <- st_coordinates(my_coords)
win <- as.owin(st_bbox(my_coords))
police_clean_ppp <- ppp(
  x = coords[,1],
  y = coords[,2],
  window = win
)
```

## Quadrat test

We want to test whether points are distributed uniformly on our surface of interest or not. To do so, we use the Quadrat test. It consists in drawing a L\*l grid around our space of concern and count the number of points included in each square of our grid. Then, a $X^2 =\Sigma_{i=1}^m\frac{(observed_i-expected)^2}{expected}$ statistic is computed, with $m$ the number of squares composing the grid, $observed_i$ the number of points contained in the rectangle $i$ and $expected = \frac{Total \space number \space of \space points}{L*l}$ the expected number of points in each square of the grid if they were to be distributed uniformly. This $X^2$ statistic converges towards a $\chi^2$. It suffices then to compare $X^2$ to the quantiles of a $\chi^2_{m-1}$ law.

There are three possible H0 for a Quadrat test :

-   If H0 : «The points are distributed uniformly in the grid» we reject H0 at level $\alpha$ if $|X^2|>\chi^2_{m-1,1-\frac{\alpha}{2}}$ . In this case, H1 is «The points are not distributed uniformly in the grid».

-   If H0 : «The points are distributed following a CSR or regular pattern in the grid» we reject HO at level $\alpha$ if $X^2>\chi^2_{m-1,1-\alpha}$. In this case, H1 is «The points are distributed following a clustered pattern in the grid»

-   If H0 «The points are distributed following a CSR or clustered pattern in the grid» we reject HO at level $\alpha$ if $X^2<\chi^2_{m-1,\alpha}$. In this case, H1 is «The points are distributed following a regular pattern in the grid».

It is important to note that the results of the test is dependent on the definition of the grid, left at the discretion of the statistician. As such, it is important to adapt the grid to the precision wanted and test different configurations to make sure we are not concluding because of statistical artifacts.

```{r}
X <- as.ppp(st_coordinates(my_coords), st_bbox(my_coords)) #Turn sf into ppp
```

We test H0 :

-   «The points are distributed uniformly in a 100\*120 grid»

-   «The points are distributed following a CSR or clustered pattern in a 100\*120 grid»

-   «The points are distributed following a CSR or regular pattern in a 100\*120 grid»

```{r}
Q <- quadratcount(X, nx = 100, ny = 120) #Transform ppp into a nx*ny grid. As England is elongated, we use an elongated grid

quadrat.test(Q) #p-value < 0.001 ==> there is a pattern of distribution
quadrat.test(Q, alternative = "regular") #p-value = 1 ==> cannot conclude on a regular pattern
quadrat.test(Q, alternative = "clustered") #p-value < 0.001 ==> clustered points around London

```

Our three tests converge towards the same answer : there is a point pattern consisting of a clustering of points. But the Quadrat method is dependent on our definition of the Q grid. Let us run the same test but with a much smaller, 2\*2 grid.\
We test H0 respectively:

-   «The points are distributed uniformly in a 2\*2 grid»

-   «The points are distributed following a CSR or clustered pattern in a 2\*2 grid»

-   «The points are distributed following a CSR or regular pattern in a 2\*2 grid»

```{r}
Q <- quadratcount(X, nx = 2, ny = 2) #Transform ppp into a nx*ny grid. As England is elongated, we use an elongated grid

quadrat.test(Q) #p-value < 0.001 ==> there is a pattern of distribution
quadrat.test(Q, alternative = "regular") #p-value = 1 ==> cannot conclude on a regular pattern
quadrat.test(Q, alternative = "clustered") #p-value < 0.001 ==> clustered points around London
```

Results remain unchanged. Let us take a much larger 1000\*1200 grid this time.

We test H0 respectively:

-   «The points are distributed uniformly in a 1000\*1200 grid»

-   «The points are distributed following a CSR or clustered pattern in a 1000\*1200 grid»

-   «The points are distributed following a CSR or regular pattern in a 1000\*1200 grid»

```{r}
Q <- quadratcount(X, nx = 1000, ny = 1200) #Transform ppp into a nx*ny grid. As England is elongated, we use an elongated grid

quadrat.test(Q) #p-value < 0.001 ==> there is a pattern of distribution
quadrat.test(Q, alternative = "regular") #p-value = 1 ==> cannot conclude on a regular pattern
quadrat.test(Q, alternative = "clustered") #p-value < 0.001 ==> clustered points around London
```

No matter the grid, it seems that London has too much points clustered in it.

Let us restrict our scope to greater London:

```{r}

#Define a bbox around London with the norm WGS84
london_bbox <- st_bbox(c(
  xmin = -0.5,
  xmax =  0.3,
  ymin = 51.3,
  ymax = 51.7
), crs = st_crs(4326))

#Conversion to sfc
london_window <- st_as_sfc(london_bbox)



my_coords <- st_as_sf(police_clean, coords = c("Longitude", "Latitude"), crs = 4326)


#Restriction to the bbox
my_coords_london <- my_coords[london_window, ]


#Conversion to ppp
X <- as.ppp(
  st_coordinates(my_coords_london),
  W = as.owin(london_bbox)
)





#Visualize the point repartition in Greater London
plot(X,size=0.1,main="Spatial repartition of the crimes in Greater London") 

```

This first plot shows that the Greater London area is very heterogeneous. The outskirts are low-density rural town with little to no crime reported. Likewise, the Thames river and the marshes appear clearly by contrast : no crimes are reported on the waters.

```{r}
#Creates a 10*10 grid used in the Quadrat test and displays the amount of crimes reported in each rectangle
Q <- quadratcount(X, nx=10, ny=10)
vals <- as.vector(as.matrix(Q))
cols <- hcl.colors(100, "YlOrRd")
col_index <- cut(vals, breaks=100, labels=FALSE)
col_matrix <- matrix(cols[col_index], nrow=nrow(Q), ncol=ncol(Q))

#Plot of Q as a heatmap
plot(Q, col=col_matrix, main="Quadrat of the crimes in Greater London")
```

This matrix shows the number of crimes reported in each square of the grid. The values that are the most extreme are the most bright ones. From this matrix, one can guess that there is obviously a concentration of points that is not random. Let us test the hypothesis same H0's as before, with a 10\*10 grid.

```{r}
Q <- quadratcount(X, nx = 10, ny = 10) #Transform ppp into a nx*ny grid. 

quadrat.test(Q) #p-value < 0.001 ==> there is a pattern of distribution
quadrat.test(Q, alternative = "regular") #p-value = 1 ==> cannot conclude on a regular pattern
quadrat.test(Q, alternative = "clustered") #p-value < 0.001 ==> clustered points around London
```

The results are no surprise : points are clustered, even when looking at Greater London only. To make sure that our results are not dependent on the definition of the grid, let us run the test on different definitions of the grid. The grids defined here are not square, to make sure that we take into account the fact that London is an ovaloid.

```{r}
Q <- quadratcount(X, nx = 150, ny = 100) #Transform ppp into a nx*ny grid. 

quadrat.test(Q) #p-value < 0.001 ==> there is a pattern of distribution
quadrat.test(Q, alternative = "regular") #p-value = 1 ==> cannot conclude on a regular pattern
quadrat.test(Q, alternative = "clustered") #p-value < 0.001 ==> clustered points around London
```

All three tests give us the same results. We do not plot the hotmap as there are too many squares in the grid. What about a much larger specification ?

```{r}
#Creates a 4*3 grid used in the Quadrat test and displays the amount of crimes reported in each rectangle
Q <- quadratcount(X, nx=4, ny=3)
vals <- as.vector(as.matrix(Q))
cols <- hcl.colors(3, "YlOrRd")
col_index <- cut(vals, breaks=3, labels=FALSE)
col_matrix <- matrix(cols[col_index], nrow=nrow(Q), ncol=ncol(Q))

#Plot of Q as a heatmap
plot(Q, col=col_matrix, main="Quadrat of the crimes in Greater London")
```

One can notice that the left and right outer layers contain far less observations than the middle ones

```{r}
Q <- quadratcount(X, nx = 4, ny = 3) #Transform ppp into a nx*ny grid.

quadrat.test(Q) #p-value < 0.001 ==> there is a pattern of distribution 
quadrat.test(Q, alternative = "regular") #p-value = 1 ==> cannot conclude on a regular pattern 
quadrat.test(Q, alternative = "clustered") #p-value < 0.001 ==> clustered points around London
```

Same as with the two grids before : the points are most definitely clustered.

# III/ Intensity estimation

The intensity function of a spatial point process $\{X_1, X_2, \dots, X_{N(A)}\}$ in a planar window $A \subset \mathbb{R}^2$ is defined as:

$$
\lambda(x) = \lim_{|dx| \to 0} \frac{\mathbb{E}[N(dx)]}{|dx|},
$$

where $dx$ is a small region containing the point $x$. Under stationary, the distribution of the process is invariant under spatial translations, and under isotropic it is invariant under rotations; therefore, the expected number of events per unit area does not depend on the location $x$, implying that the intensity is constant and equal to $\lambda(x) = \lambda = \frac{\mathbb{E}[N(A)]}{|A|}$.

Thus, for an observed spatial point pattern of $n$ events observed in a region $A$, the intensity can be estimated as:

$$
\hat{\lambda} = \frac{n}{|A|}.
$$

The density and intensity functions are proportional:

$$
\lambda(x) = f(x) \int_A \lambda(u)\, du,
$$

where $\int_A \lambda(u)\, du$ is the expected number of events in $A$. So the intensity is equal to the probability to observe an event at location $x$ multiply by the expected number of event in $A$

$$
\hat{f}(x) = \frac{1}{n} \sum_{i=1}^{n} \frac{1}{h^2} 
K\!\left( \frac{x - x_i}{h} \right).
$$

$$
\hat{\lambda}(x) = \sum_{i=1}^{n} \frac{1}{h^2} 
K\!\left( \frac{x - x_i}{h} \right).
$$

where $K(\cdot)$ is a symmetric function such that $K(x) \ge 0 \quad \forall x$ and $\int_A K(x)\, dx = 1,$ and $h$ is a smoothing parameter known as the bandwidth.

where $K(\cdot)$ is a symmetric function such that $K(x) \ge 0 \quad \forall x$ and $\int_{\mathbb{R}^2} K(u)\, du = 1$.

If a Gaussian kernel is used in two dimensions, then

$$
K(u) = \frac{1}{2\pi} 
\exp\left(-\frac{1}{2} u^\top u \right),
\quad u \in \mathbb{R}^2,
$$

where $u^\top u = \|u\|^2 = u_1^2 + u_2^2$.

One edge-correction term is defined as:

$$
p_h(x) = \int_A h^{-2} 
K\!\left( \frac{x - u}{h} \right) du
$$

$p_h(x)$ represents the proportion of the kernel centered at $x$ that lies within the study region $A$ to correct the negative bias that occurs near the boundaries.

$$ 
\longrightarrow \hat{f}(x) = \frac{1}{n p_h(x)}  \sum_{i=1}^{n} \frac{1}{h^2} 
K\!\left( \frac{x - x_i}{h} \right)
$$

and

$$
\longrightarrow \hat{\lambda}(x) = \frac{1}{p_h(x)}  \sum_{i=1}^{n} \frac{1}{h^2} 
K\!\left( \frac{x - x_i}{h} \right)
$$

```{r}

my_coords2 <- st_as_sf(police_clean, coords = c("Longitude", "Latitude"), crs = 4326)


my_coords2 = st_transform(my_coords2, crs = 27700)
# Coordinates
coords2 <- st_coordinates(my_coords2)



map <- ne_countries(type = "countries", country = "United Kingdom",
                    scale = "small", returnclass = "sf")


map <- st_transform(map, crs = "EPSG:27700")
win <- as.owin(map)


pp <- ppp(
  x = coords2[,1],
  y = coords2[,2],
  window = win
)


par(mfrow = c(1,1),
    mar = c(1,1,1,1))


# Plot
plot(win, col = "lightgrey", main = "Spatial Point Pattern of Police Outcomes in the UK")
plot(pp, add = TRUE, pch = 20, col = "blue",size = 0.8)






```

## Intensity estimation

Possible kernels are : gaussian, epanechnikov, quartic, disc or a personalized function

Estimation with the default bandwidth and gaussian kernel:

```{r}
# kernel gaussian
lambdahat <- density(pp)
attr(lambdahat, "sigma")
```

Interpretation:

Sigma = bandwidth, it is in meters so a point has an influence on intensity in a zone with 76 km of diameter

```{r}
par(mfrow = c(2,2),
    mar = c(1,1,1,1),
    oma = c(0,0,2,0))

plot(density(pp, sigma = 10000),
     main = "Bandwidth = 10000")
contour(density(pp, sigma = 10000), add = TRUE) # contour plot

plot(density(pp, sigma = 20000),
     main = "Bandwidth = 20000")
contour(density(pp, sigma = 20000), add = TRUE) # contour plot



plot(density(pp, sigma = 50000),
     main = "Bandwidth = 50000")
contour(density(pp, sigma = 50000), add = TRUE) # contour plot



plot(density(pp),
     main = "Default Bandwidth = 76634.26")
contour(density(pp), add = TRUE) # contour plot


# Titre global
mtext("Intensity estimation with different bandwidths",
      outer = TRUE,
      cex = 1.5,
      font = 2)


```

### Focus on London

```{r}
# Define bbox of London (EN WGS84 !)
london_bbox <- st_bbox(c(
  xmin = -0.6,
  xmax =  0.3,
  ymin = 51.3,
  ymax = 51.7
), crs = 4326)

# Transform bbox in 27700
london_window <- st_as_sfc(london_bbox)
london_window <- st_transform(london_window, crs = 27700)

# Filtrer points in London

my_coords3 = my_coords2 # |> subset(Outcome.type == "Suspect charged")

my_coords_london <- st_intersection(my_coords3, london_window)

# Coordinates
coords_london <- st_coordinates(my_coords_london)

# Spatial Window (IMPORTANT : use london_window instead of bbox)
win_london <- as.owin(london_window)

# Generate ppp
pp_london <- ppp(
  x = coords_london[,1],
  y = coords_london[,2],
  window = win_london
)

# Plot
par(mfrow = c(1,1), mar = c(1,1,1,1))

plot(win_london, col = "lightgrey", main = "Police Outcomes in London")
plot(pp_london, add = TRUE, pch = 20,size = 0.1, col = "blue")


```

```{r}
# kernel gaussian
lambdahat <- density(pp_london)
attr(lambdahat, "sigma")
```

```{r, fig.width=10, fig.height=10}


plot(density(pp_london, sigma = 1000),
     main = "Bandwidth = 1000")
contour(density(pp_london, sigma = 1000), add = TRUE) # contour plot


plot(density(pp_london, sigma = 1250),
     main = "Bandwidth = 1250")
contour(density(pp_london, sigma = 1250), add = TRUE) # contour plot




plot(density(pp_london, sigma = 500),
     main = "Bandwidth = 500")
contour(density(pp_london, sigma = 500), add = TRUE) # contour plot


plot(density(pp_london, sigma = 1500),
     main = "Bandwidth = 1500")
contour(density(pp_london, sigma = 1500), add = TRUE, log = TRUE) # contour plot



```

We notice a hotspot of crime in center of London

```{r,fig.width=6, fig.height=10}



# Data in sf + projection UK
my_coords2 <- st_as_sf(police_clean, coords = c("Longitude", "Latitude"), crs = 4326)
my_coords2 <- st_transform(my_coords2, crs = 27700)


my_coords3 = my_coords2 # |> subset(Outcome.type == "Suspect charged")


# BBOX London (WGS84)
london_bbox <- st_bbox(c(
  xmin = -0.6,
  xmax =  0.3,
  ymin = 51.3,
  ymax = 51.7
), crs = 4326)

#  Conversion bbox + projection 27700
london_window <- st_as_sfc(london_bbox)
london_window <- st_transform(london_window, crs = 27700)

#  Filtrer points dans Londres
my_coords_london <- st_intersection(my_coords3, london_window)

#  Coordonnées
coords_london <- st_coordinates(my_coords_london)

#  Fenêtre spatiale
win_london <- as.owin(london_window)

# Objet ppp
pp_london <- ppp(
  x = coords_london[,1],
  y = coords_london[,2],
  window = win_london
)




# Transformation des points et bbox pour leaflet
points_wgs84 <- st_transform(my_coords_london, 4326)
london_window_wgs84 <- st_transform(london_window, 4326)


# all geographic unit from UK
LSOA_uk = st_read("LSOA.gpkg") 
LSOA_uk <- st_transform(LSOA_uk, 4326)





# london_window_wgs84 = Window polygone

# Found LSOA that touch london_window_wgs84

index <- st_intersects(LSOA_uk, london_window_wgs84)

# We keep LSOA that at least one intersection with london_window_wgs84
LSOA_in_window <- LSOA_uk[lengths(index) > 0, ]



london_full =st_union(LSOA_in_window) 


# Coherent Window
london_full_win <- as.owin(st_transform(london_full, 27700))


# ppp object
pp_london <- ppp(
  x = coords_london[,1],
  y = coords_london[,2],
  window = london_full_win
)



#  Kernel density
dens <- density(pp_london, sigma = 3000)

#  Transform to raster

r <- rast(dens)
crs(r) <- "EPSG:27700"

# Projection
r_wgs84 <- project(r, "EPSG:4326")


# Colors
pal <- colorNumeric("viridis", dens$v, na.color = "transparent")

#  Leaflet map
leaflet() %>%
  addTiles() %>%
  
  # Heatmap (densité)
#  addRasterImage(
#    r_wgs84,
#    colors = pal,
#    opacity = 0.6
#  ) %>%
  
  # Contour Londres
  #addPolygons(
  #  data = london_full,
  #  fill = FALSE,
  #  color = "black",
  #  weight = 2
 # ) %>%
  
  # Points
  addCircleMarkers(
    data = points_wgs84,
    radius = 1,
    color = "black",
    opacity = 0.2,
    stroke = FALSE
  ) |> addTiles("Spatial Point Pattern of crimes in London")

```

### Import of population table (variable aggregated at LSOA level)

LSOA = geographic unit used by ONS (organism national statistics) of UK

source: [United kingdom government: population estimate by LSOA](https://www.ons.gov.uk/peoplepopulationandcommunity/populationandmigration/populationestimates/datasets/lowersuperoutputareamidyearpopulationestimates)

LSOA table : <https://geoportal.statistics.gov.uk/datasets/0f80c523f3cd4d0fab5111572f84a2fb_0/explore>

Plot of all geographic unit

```{r}



leaflet(LSOA_in_window) %>%
  addTiles() %>%
  addPolygons(
    weight = 0.5,
    color = "blue",
    fill = FALSE,
    opacity = 1  
    ) |>
  addPolygons(
  data = london_full,
  fill = FALSE,
  color = "black",
  weight = 2
  ) |>  addTiles("Area with population information in London")



```

```{r}

population <- read_excel("population.xlsx", 
    sheet = "Mid-2024 LSOA 2021", skip = 3)
population = population[,1:5]

```

```{r}

LSOA_in_window |> colnames()
LSOA_in_window_data_frame = LSOA_in_window[,1:4] |> as.data.frame()

pop_london = LSOA_in_window |> left_join(population, by = c("LSOA01CD" = "LSOA 2021 Code"))


sum(is.na(pop_london$Total))

```

We remplace NA by mean for 3 nearest regions

```{r}


# Check pop_london is sf
pop_london_sf <- st_as_sf(pop_london)

# Check for NAs
sum(is.na(pop_london_sf$Total))

# Separate NA and non-NA
na_rows <- pop_london_sf %>% filter(is.na(Total))
non_na_rows <- pop_london_sf %>% filter(!is.na(Total))

# Loop to replace NAs by the mean of the three closest neighbors
for(i in seq_len(nrow(na_rows))) {
  
  # Centroid of object
  centroid_i <- st_centroid(na_rows[i, "SHAPE"])
  
  # Compute distance to non-NA objects
  dists <- st_distance(centroid_i, st_centroid(non_na_rows$SHAPE))
  
  # Sort and take 3-NN closest
  nearest_idx <- order(dists)[1:3]
  
  # Mean neighbors
  na_rows$Total[i] <- mean(non_na_rows$Total[nearest_idx], na.rm = TRUE)
}

# bind rows
pop_london_imputed <- bind_rows(non_na_rows, na_rows)

# Check no more NAs
sum(is.na(pop_london_imputed$Total))


```

We transform pop_london_sf in a point patern weighted population by taking centroid of each region

```{r}
# Compute centroids
pop_london_points <- st_centroid(pop_london_imputed)

radius <- rescale(sqrt(pop_london_points$Total), to = c(1, 6))

leaflet(pop_london_points) %>%
  addTiles() %>%
  addCircleMarkers(
    radius = radius,
    stroke = FALSE,
    fillOpacity = 0.6,
    color = "red",
    popup = ~paste0("Population: ", Total)
  )
```

Plot of intensity population in London

```{r, fig.height=10, fig.width=10}

# Reprojection before centroid
pop_london_proj <- st_transform(pop_london_imputed, 27700)

# Centroids in the right system
pop_london_points <- st_centroid(pop_london_proj)

# Coherent coordinates
coords <- st_coordinates(pop_london_points)

# Coherent window
win <- as.owin(st_union(pop_london_proj))

# Weighted point pattern
pop_point_pattern <- ppp(
  x = coords[,1],
  y = coords[,2],
  window = win,
  marks = pop_london_points$Total
)

# Weighted estimation of the Kernel
dens <- density(
  pop_point_pattern,
  weights = pop_point_pattern$marks,
  sigma = bw.diggle(pop_point_pattern)
)
# Plot
plot(dens, main = "Intensity population (kernel)")
contour(dens, add = TRUE)

```

```{r}


# Tranform into bbox for leaflet
points_wgs84 <- st_transform(my_coords_london, 4326)
london_window_wgs84 <- st_transform(london_window, 4326)


# Convert to raster
dens_raster <- rast(dens)

crs(dens_raster) <- "EPSG:27700"

# Projection in WGS84 for leaflet
# Projection
r_wgs84 <- project(dens_raster, "EPSG:4326")
# Colors
pal <- colorNumeric("viridis", values(r_wgs84), na.color = "transparent")

# Plot leaflet
leaflet() %>%
  addTiles() %>%
  addRasterImage(
    r_wgs84,
    colors = pal,
    opacity = 0.8
  ) 


```

### Intensity ratio

Intensity Ratio and Spatial Relative Risk

We consider two spatial point patterns observed over the same region: a set of cases (for example, individuals affected by a disease) and a set of controls representing the population at risk. The objective is to compare these two patterns in order to identify spatial variations in risk and detect potential high-risk areas.

We assume that the cases arise from a Poisson process with intensity $\lambda(x)$, while the controls arise from an independent Poisson process with intensity $\lambda_0(x)$. The relationship between these two intensities can be expressed as: $$
\lambda(x) = \alpha \, \lambda_0(x) \, \rho(x)
$$

In this decomposition, $\rho(x)$ represents the spatial relative risk, while $\alpha$ is a scaling factor introduced to account for the difference in the number of observed points between the two patterns.

Since the number of controls is typically larger than the number of cases, the scaling factor is estimated as:

$$
\hat{\alpha} = \frac{n_{\text{cases}}}{n_{\text{controls}}}
$$

This correction ensures that the comparison between the two intensities is not biased by differences in sample size.

The intensity functions $\lambda(x)$ and $\lambda_0(x)$ are not known and must be estimated from the data. This is commonly done using kernel smoothing methods. The estimator of the intensity at location $x$ is given by:

$$
\hat{\lambda}(x) = \sum_{i=1}^{n} K_\sigma(x - x_i)
$$

where $K_\sigma$ is a kernel function (typically Gaussian) and $\sigma$ is the bandwidth parameter controlling the degree of smoothing. In practice, we compute two estimates: $\hat{\lambda}(x)$ for the cases and $\hat{\lambda}_0(x)$ for the controls.

The spatial relative risk is then estimated by taking the ratio of these smoothed intensities:

$$
\hat{\rho}(x) = \frac{\hat{\lambda}(x)}{\hat{\alpha} \, \hat{\lambda}_0(x)}
$$

This estimator provides a continuous surface describing how the risk varies across space.

An important practical issue is the choice of the bandwidth. To ensure a fair comparison between the two intensity estimates, it is recommended to use a common bandwidth for both datasets. A simple and commonly used approach consists in taking the average of the optimal bandwidths computed separately for cases and controls:

$$
bw = \frac{bw_{\text{cases}} + bw_{\text{controls}}}{2}
$$

The interpretation of the relative risk surface is straightforward. When $\hat{\rho}(x) > 1$, the location $x$ corresponds to an area where the risk is higher than average, often referred to as a hotspot. When $\hat{\rho}(x) < 1$, the risk is lower than average. Values close to 1 indicate no significant deviation from the baseline risk.

Finally, the estimated relative risk surface is typically visualized as a map, which allows for an intuitive identification of spatial patterns. This approach is widely used in spatial epidemiology, although it remains sensitive to the choice of bandwidth and relies on the assumption of Poisson processes.

Crime VS population

```{r}
# crimes = ppp_crimes
# population = pop_point_pattern (avec marks = Total)

# 1. Generate ppp population not marker (only positions)
pop_ppp <- unmark(pop_point_pattern)

# 2. Common pass-through band
#bw_crime <- attr(density(pp_london), "sigma")
bw_crime = 1000
bw_pop <- attr(density(pop_ppp), "sigma")
bw <- (bw_crime + bw_pop) / 2

# 3. Densities
int_crime <- density(pp_london, sigma = bw)
int_pop <- density(pop_ppp, sigma = bw, weights = pop_point_pattern$marks)

# 4. alpha
alpha <- pp_london$n / sum(pop_point_pattern$marks)

# 5. Ratio
rr <- t(int_crime$v) / t(alpha * int_pop$v)

# 6. Plot
image.plot(int_crime$xcol, int_crime$yrow, rr, asp = 1,
           main = "Relative Risk (Crimes charged vs Population)")
#contour(int_crime, add = TRUE)

```

In conclusion, even after controlling for the population at risk, the analysis shows that the center of London exhibits a substantially higher relative risk, indicating that individuals in this area are much more likely to be affected.

# IV/ The k-function

The K-function is defined as :

$$
K(s) = \frac{E(\text{number of n events within distance s of event i})}{\lambda}
$$ Where $\lambda$ is the intensity function. As mentioned before, it is defined, in the case of CSR as:

$$
\lambda = \frac{E(N(A))}{|A|}
$$ Where N(A) is the number of points in A and $|A|$ is the area of A.

We can also rewrite the "number of n events within distance s of event i" as the number of points within a circle C of radius s and center i or more concise within region A: "number of points in C". Then, we can rewrite the K-function as:

$$
K(s) = \frac{E(\text{number of points in C})\times\frac{|C|}{|C|}}{\frac{E(\text{number of points in C)} }{|C|}}
$$ Which simplifies to : $K(s) = |C|=\pi\times s^2$.

As we mentioned before, CSR is our benchmark: this scenario means that all points are distributed randomly and there is no clustering or regular patterns. Meaning that points are neither too close or too separated. This interpretation is due to the fact that if we had a clustering situation while we suppose CSR, then it would be highly likely that there will be more points than expected under CSR close to a certain event i, thus driving the expectation in the numerator above the CSR scenario. In the case of regular pattern, we would have the opposite situation, there would be points that are quite separeted driving the number of expected points arround a certain event i down and thus decreasing K.

Before going to the application, let's define the estimator function of K when we do not assume CSR:

$$
\hat{K}(s) = \frac{\hat{E}(s)}{\hat{\lambda}}
$$ Where: $$
\hat{E}(s)=\frac{1}{n}\sum_{i=1}^n\sum_{i \neq j}^{n-1}w_{ij}I(d_{ij}\le s)
$$ and $\hat{\lambda}$ the estimator of the intensity function as defined before.

This estimator is a quite forward estimation, in fact for the numerator we just count the number of points we actually see in the circle C for all possible points and we then take the mean to retrieve the average number of points in the circle C when we take different centroids. This gives us a clear idea of the number of neighboring events for each point on average. For the denominator, we just take the usual estimator for the intensity function as we saw before.

About the weights, we need them because there are certain events that cannot be considered as the others. For example, if we consider an event i that is at border of A, if we do not take into consideration the fact that a part of the circle C is located outside the region and thus has 0 probability of containing other events, we would have a negative bias for these points. Another reason to consider this weights is because: we might be considering a circle C that is too small or too large: in fact if we take a circle so small that cannot contain any other event, then we would have a negative bias. Inversely, if the circle is too big, we might be considering as neighbors events that are too far apart, thus producing a positive bias.

The weights we can apply in the package spatstat are:

-   The border correction (Ripley 1988): $w_{ij}=1$ if C fits A, else it is equal to 0 (Ripley, 1988)

-   The isotropic correction (Ohser 1983, Ripley 1988): $w_{ij} = \frac{\text{Total area of C}}{\text{Are of C in A}}$

-   The translate correction (Ohser 1983): $$
    w_{ij} = \frac{|A|}{|A \cap (A+(x_i-x_j))|}
    $$ It exist also the possibility to compute $\hat{K}(s)$ without $w_{i,j}$.

```{r}
K <- Kest(
  police_clean_ppp,
  correction = c("border", "Ripley", "isotropic", "translate")
)
plot(K)
```

The function K is giving us the mean of all possible circles of size r for each r between 0 and 25 km. We clearly see that we are above the benchmark of the theoretical K (when we apply CSR) whatoever the weights, then we can conclude that points are clearly clustered.

### The L-function

A variant of the $\hat{K}(s)$ function is the $\hat{L}(s)$ that is the square root of the first one divided by square root of $\pi$. This variant allows to interpret better the measure of the K-function and gives us better visualizations.

$$
L(s) = \sqrt{\frac{K(s)}{\pi}}
$$ and $\hat{L}(s) = \sqrt{\frac{\hat{K}(s)}{\pi}}$

```{r}
L <- Lest(police_clean_ppp,
          correction = c("border", "Ripley", "isotropic", "translate")
)
plot(L)
```

We have the same interpretations for the L-function.

### The K-function and Monte Carlo

It is also possible to test for CSR using the K function and a Monte Carlo approximation:

-   Compute the $\hat{K}(s)$ of the real data.
-   Generate M number of spatial point patterns over A using homogeneous Poisson process (CSR).
-   Estimate $\hat{K}(s)$ and take the confidence interval of size 95%.
-   Reject CSR if the real estimation is outside of the CI.

This procedures gives us a more significant measure to test for CSR. In fact, it allows us (or not) to reject at confidence level 95% the precense of CSR.

```{r}
E <- envelope(police_clean_ppp, Kest, nsim = 99)
plot(E)
```

### Conclusion for K-function

Clearly the pattern is not random as the poisson distribution is clearly lower than the empirical K. The estimators (with the different corrections) are quite high meaning that the point pattern is clearly not random. For small radius we observe that the clustering is already important but it clearly explodes after 5 km.

Remark: one limitation, as we could see on the map above is that the points are quite centered in London and surroundings meaning that our estimation is driven mainly by this area.

# V/ Regression

### The Model We Are Using

We want to understand where and why police incidents occur across different cities in the UK. Instead of assuming that incidents happen completely at random, we model their spatial distribution as a function of local characteristics: population and GDP/Habitant.

#### Model Specification

We use an Inhomogeneous Poisson Point Process Model. This is a model that allows the intensity of events (police incidents) to vary across space depending on covariates such as the number of residents.

The expected intensity of incidents at any location $u$ is given by:

$$
\lambda(u) = \exp(\beta_0 + \beta \cdot X(u))
$$

where: $\lambda(u)$ is the intensity (expected number of incidents per unit area) at location $u$, $X(u)$ is a characteristic in the area containing location $u$ (like population or GDP/Habitant), $\beta_0$ is the intercept, $\beta$ measures the effect of population on the intensity of incidents.

A positive $\beta$ would indicate that areas with more of a certain characteristic (more residents or higher GDP/Habitant) tend to have a higher intensity of police incidents, while a negative value suggests the opposite.

#### Key Assumptions

-   The points are generated by an inhomogeneous Poisson process.
-   Conditional on the covariates, the occurrence of events at different locations is independent.

### Application 

We will do regression using the dataset from this website: <https://www.ons.gov.uk/economy/grossdomesticproductgdp/datasets/regionalgrossdomesticproductlocalauthorities?utm_source=chatgpt.com> It contains diverse indicators on areas in the United Kingdom, it was published in December 2019.

First we need to import the shapes of the districts in the UK

```{r}
areas <- st_read("Shape/Local_Authority_Districts_December_2019_GCB_UK.shp")
```

```{r}
areas |>
  dplyr::select(lad19cd, lad19nm) |>
  head() #It works
```

We will use two variables:

-   Total resident population numbers

-   Gross domestic product per head at current market prices

## For resident population numbers

```{r}
path_data = paste(getwd(), "regionalgrossdomesticproductgdplocalauthorities.xlsx", sep = "/")

number_residents <- read_excel(
  path_data,
  sheet = "Table 6",
  skip = 1
)
```

Renaming so that population and shapes coincide

```{r}
areas <- areas |> dplyr::rename( 
  area_code = "lad19cd", 
  area_name = "lad19nm" 
  )

number_residents <- number_residents |> dplyr::rename( 
  area_code = "LA code", 
  area_name = "LA name" 
  )

areas <- areas |>
  dplyr::left_join(number_residents, by = "area_code")

areas_vect <- as(areas, "Spatial")
```

Plot it

```{r}
plot(areas["2018"])
```

London area population/crimes:

```{r}
london_area <- areas %>% subset(`ITL1 Region` == 'London')
london_crimes <- my_coords[grepl("^E0100", my_coords$LSOA.code), ]
mapview(london_area["2018"]) +
  mapview(london_crimes, col.regions = "red", alpha = 0.2, cex = 0.5)
```

It seems that there is no major link between the crimes commited and the number of residents in an area. We will try regression anyways.

Proceed to regression

```{r}
#A raster is a database organized as a rectangular grid that is sub-divided into rectangular cells of equal area (in terms of the units of the coordinate reference system)
r_template <- raster(extent(areas_vect))  #we use extent to get the same size as the number of points in the police dataset
population_raster <- rasterize(
  areas_vect,          #Our polygones
  r_template,          #Raster base where they will be put
  field = "X2018",     #Population in 2018
  fun = mean,          #Take mean if various polygones are in the same area
  background = NA      #NA for cases without any value
)

#Transform ppp into spatial points
points_sp <- SpatialPoints(cbind(police_clean_ppp$x, police_clean_ppp$y),
                           proj4string = crs(population_raster))

#Extract population values for each point
police_clean_ppp$number_residents <- raster::extract(population_raster, points_sp)

#Only keep points where the value of population is not NA
police_clean_ppp_filtered <- police_clean_ppp[!is.na(police_clean_ppp$number_residents)]
population_raster[is.na(population_raster)] <- 0 #To double check (give them value 0)

#Convert to im so that we can use it in ppm
mat <- as.matrix(population_raster)
ext <- extent(population_raster)
number_residents_im <- im(
  mat,
  xcol = seq(ext@xmin, ext@xmax, length.out = ncol(mat)),
  yrow = seq(ext@ymin, ext@ymax, length.out = nrow(mat))
) 
```

Result

```{r}
ppm_model <- ppm(
  police_clean_ppp ~ number_residents,
  covariates = list(number_residents = number_residents_im)
)
summary(ppm_model)
```

Model 1:

Notes:

-   Edge correction "border"

-   82075 points

-   Average intensity: 4.51e-07

-   Area of the window: 1.81916e+11 square units

-   Dummy points: fictive points used to approximate integral

-   Intercept: -14.41 in logs (base log-intensity)

-   When we increase by one unit the population, the log intensity deacreases by -5.118290e-05.

    or alternatively, when we increase by 1 unit population, intensity is multiplied by 0.9999.

What is probably happening is that the number of inhabitants in a region might not explain by itself the number of crimes, it is probably a problem of IV.

## For product per head

```{r}
product_head <- read_excel(
  path_data,
  sheet = "Table 7",
  skip = 1
)

areas <- st_read("Shape/Local_Authority_Districts_December_2019_GCB_UK.shp")

areas |>
  dplyr::select(lad19cd, lad19nm) |>
  head() #It works

areas <- areas |> dplyr::rename( 
  area_code = "lad19cd", 
  area_name = "lad19nm" 
  ) 

product_head <- product_head |> dplyr::rename( 
  area_code = "LA code", 
  area_name = "LA name" 
  )

areas <- areas |>
  dplyr::left_join(product_head, by = "area_code")

areas$`2023` <- as.numeric(areas$`2023`)

areas_vect <- as(areas, "Spatial")

plot(areas["2023"])
```

```{r}
london_area <- areas %>% subset(`ITL1 Region` == 'London')
london_area <- london_area %>%
    dplyr::filter(!(area_code %in% c("E09000001", "E09000007"))) #Take out observations with +1M GDP/head : City of London and Camdem
london_crimes <- my_coords[grepl("^E0100", my_coords$LSOA.code), ]
mapview(london_area["2023"]) +
  mapview(london_crimes, col.regions = "red", alpha = 0.2, cex = 0.5)
```

```{r}
#A raster is a database organized as a rectangular grid that is sub-divided into rectangular cells of equal area (in terms of the units of the coordinate reference system)
r_template <- raster(extent(areas_vect))  #we use extent to get the same size as the number of points in the police dataset
gdphead_raster <- rasterize(
  areas_vect,          #Our polygones
  r_template,          #Raster base where they will be put
  field = "X2023",     #GDP/head in 2023
  fun = mean,          #Take mean if various polygones are in the same area
  background = NA      #NA for cases without any value
)

#Transform ppp into spatial points
points_sp <- SpatialPoints(cbind(police_clean_ppp$x, police_clean_ppp$y),
                           proj4string = crs(gdphead_raster))

#Extract GDP/head values for each point
police_clean_ppp$gdp_hab <- raster::extract(gdphead_raster, points_sp)

#Only keep points where the value of GDP/head is not NA
police_clean_ppp_filtered <- police_clean_ppp[!is.na(police_clean_ppp$gdp_hab)]
gdphead_raster[is.na(gdphead_raster)] <- 0 #To double check (give them value 0)

#Convert to im so that we can use it in ppm
mat <- as.matrix(gdphead_raster)
ext <- extent(gdphead_raster)
gdp_hab_im <- im(
  mat,
  xcol = seq(ext@xmin, ext@xmax, length.out = ncol(mat)),
  yrow = seq(ext@ymin, ext@ymax, length.out = nrow(mat))
) 
```

```{r}
ppm_model <- ppm(
  police_clean_ppp ~ gdp_hab,
  covariates = list(gdp_hab = gdp_hab_im)
)
summary(ppm_model)
```

Model 2:

-   Intercept: -14.41 in logs (base log-intensity)

-   When we increase by one unit the GDP/capita, the log intensity deacreases by -2.654507e-04

    or alternatively, when we increase by 1 unit population, intensity is multiplied by 0.9997.

## Final regression

```{r}
ppm_model <- ppm(
  police_clean_ppp ~ gdp_hab + number_residents,
  covariates = list(gdp_hab = gdp_hab_im, number_residents = number_residents_im)
)
summary(ppm_model)
```

Model 3:

-   For each extra unit of GDP/habitant, the intensity of crime is reduced by -2.654507e-04

-   For each time there is an increase of one unit of GDP/habitant, we multiply the intensity of crime by 0.9995 (or reduce it by 5%)

-   For the population, when increasing by one unit the number of residents, the intensity of crime is increased by 2.561542e-05.

-   For each time there is an increase of one unit of population, we multiply the intensity of crime by 1.00002 (really close to 1).

-   Overall, even if the two coefficients are significatives, the estimate for population is quite small. There is however a clear impact of GDP/capita on the intensity of crime.

-   The fact that we saw a change in the sign of the population estimate when going from model 1 to model 3 indicates that there might be a problem of endogeneity that should be studied with IV for example.

# Conclusion

The three methods to test spatial correlation presented conclude that there is a strong correlation between the crimes and the regions. Crimes do not happen randomly at a given place but they are driven by certain factors. We regressed the crimes on the population and on gdp per capita separetly and we find that there is a clear link between the GDP per capita and the crimes. The higher the GDP per capita the lower the crimes. When regressing the two variables together GDP per capita decreases the number of crimes and the number of habitants increases it. However, the coefficients are small. This indicates that there might be some other cofunders and perhaps the use of IV might be useful.
